---
title: "F1 Pit Strategy Neural Network"
author: "Blake Thompson and Cameron"
date: "4/27/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Project goal

This project trains a simple neural network to predict whether a Formula 1 driver will pit on the next lap.

The target column is `PitNextLap`:

- `0` means the driver will not pit on the next lap.
- `1` means the driver will pit on the next lap.

The main idea is: use the current race state, tire state, driver, race, and lap timing information to predict the next-lap pit decision.

## Load packages

```{r}
library(keras3)
library(tensorflow)
```

## Load the dataset

The dataset should stay in the `data` folder.

```{r}
set.seed(42)

data_path <- file.path("data", "f1_strategy_dataset_v4.csv")
f1df <- read.csv(data_path, stringsAsFactors = FALSE, check.names = FALSE)

summary(f1df)
head(f1df)
```

## Look at the target variable

Before training, check whether the label is balanced. For this project, `PitNextLap = 1` is much rarer than `PitNextLap = 0`, so accuracy alone can be misleading.

```{r}
table(f1df$PitNextLap)
prop.table(table(f1df$PitNextLap))
```

## Choose the columns used by the model

`PitNextLap` is the label, so it cannot be used as an input feature.

`PitStop` is also removed because it is too close to the answer. If the model sees information that directly tells it whether a pit event happened, then the model is not really learning a useful prediction pattern.

```{r}
label_column <- "PitNextLap"

categorical_features <- c("Driver", "Compound", "Race")

numeric_features <- c(
  "LapNumber",
  "Stint",
  "TyreLife",
  "Position",
  "LapTime (s)",
  "Year",
  "LapTime_Delta",
  "Cumulative_Degradation",
  "RaceProgress",
  "Normalized_TyreLife",
  "Position_Change"
)

feature_columns <- c(categorical_features, numeric_features)

# Check that the dataset has every column this file expects.
required_columns <- c(feature_columns, label_column, "PitStop")
missing_columns <- setdiff(required_columns, names(f1df))

if (length(missing_columns) > 0) {
  stop(paste("Missing columns:", paste(missing_columns, collapse = ", ")))
}
```

## Optional quick mode

Set `quick_mode <- TRUE` while testing the file. This trains on fewer race-year groups so the R Markdown file runs faster.

Set `quick_mode <- FALSE` for the more complete final run.

```{r}
quick_mode <- TRUE
max_quick_groups <- 24

f1df$race_year_group <- paste(f1df$Race, f1df$Year, sep = "_")

if (quick_mode) {
  all_groups <- unique(f1df$race_year_group)
  picked_groups <- sample(all_groups, size = min(max_quick_groups, length(all_groups)))
  f1df <- f1df[f1df$race_year_group %in% picked_groups, ]
}

nrow(f1df)
length(unique(f1df$race_year_group))
```

## Split data into training, validation, and testing

The split is done by `Race + Year`, not by individual rows.

That matters because rows from the same race are related. If the same race appears in both training and testing, the test score can look better than it really is.

```{r}
set.seed(42)

groups <- unique(f1df$race_year_group)
groups <- sample(groups)

if (length(groups) < 3) {
  stop("Need at least three race-year groups to make train, validation, and test splits.")
}

n_groups <- length(groups)
train_count <- max(1, floor(n_groups * 0.70))
validation_count <- max(1, floor(n_groups * 0.15))

# Always leave at least one race-year group for validation and one for testing.
if (train_count + validation_count >= n_groups) {
  validation_count <- 1
  train_count <- n_groups - validation_count - 1
}

train_groups <- groups[seq_len(train_count)]
validation_groups <- groups[(train_count + 1):(train_count + validation_count)]
test_groups <- groups[(train_count + validation_count + 1):n_groups]

training <- f1df[f1df$race_year_group %in% train_groups, ]
validation <- f1df[f1df$race_year_group %in% validation_groups, ]
testing <- f1df[f1df$race_year_group %in% test_groups, ]

split_summary <- data.frame(
  split = c("training", "validation", "testing"),
  rows = c(nrow(training), nrow(validation), nrow(testing)),
  pit_next_lap_rate = c(
    mean(training$PitNextLap),
    mean(validation$PitNextLap),
    mean(testing$PitNextLap)
  )
)

split_summary
```

## Data preprocessing

The neural network needs numbers, not raw strings.

So this section does three things:

1. Missing numeric values are replaced with the training median.
2. Numeric columns are scaled using the training mean and standard deviation.
3. Categorical columns are one-hot encoded.

The important rule is: preprocessing values are learned from the training set only. Validation and testing reuse the training preprocessing rules.

```{r}
mode_value <- function(values) {
  values <- values[!is.na(values) & values != ""]
  if (length(values) == 0) {
    return("UNKNOWN")
  }
  names(sort(table(values), decreasing = TRUE))[1]
}

fit_preprocessor <- function(train_df) {
  category_fill_values <- list()
  category_levels <- list()
  
  for (feature in categorical_features) {
    if (feature == "Compound") {
      fill_value <- "UNKNOWN"
    } else {
      fill_value <- mode_value(train_df[[feature]])
    }
    values <- as.character(train_df[[feature]])
    values[is.na(values) | values == ""] <- fill_value
    
    category_fill_values[[feature]] <- fill_value
    category_levels[[feature]] <- sort(unique(c(values, fill_value, "OTHER")))
  }
  
  numeric_medians <- c()
  numeric_means <- c()
  numeric_sds <- c()
  
  for (feature in numeric_features) {
    values <- as.numeric(train_df[[feature]])
    median_value <- median(values, na.rm = TRUE)
    values[is.na(values)] <- median_value
    
    mean_value <- mean(values)
    sd_value <- sd(values)
    
    if (is.na(sd_value) || sd_value == 0) {
      sd_value <- 1
    }
    
    numeric_medians[[feature]] <- median_value
    numeric_means[[feature]] <- mean_value
    numeric_sds[[feature]] <- sd_value
  }
  
  list(
    category_fill_values = category_fill_values,
    category_levels = category_levels,
    numeric_medians = numeric_medians,
    numeric_means = numeric_means,
    numeric_sds = numeric_sds
  )
}

transform_data <- function(input_df, preprocessor) {
  # Build numeric matrix.
  x_numeric <- matrix(
    0,
    nrow = nrow(input_df),
    ncol = length(numeric_features)
  )
  colnames(x_numeric) <- numeric_features
  
  for (feature in numeric_features) {
    values <- as.numeric(input_df[[feature]])
    values[is.na(values)] <- preprocessor$numeric_medians[[feature]]
    
    x_numeric[, feature] <- (
      values - preprocessor$numeric_means[[feature]]
    ) / preprocessor$numeric_sds[[feature]]
  }
  
  # Build categorical dataframe, then one-hot encode it.
  categorical_df <- data.frame(row_id = seq_len(nrow(input_df)))
  
  for (feature in categorical_features) {
    values <- as.character(input_df[[feature]])
    fill_value <- preprocessor$category_fill_values[[feature]]
    levels <- preprocessor$category_levels[[feature]]
    
    values[is.na(values) | values == ""] <- fill_value
    values[!values %in% levels] <- "OTHER"
    
    categorical_df[[feature]] <- factor(values, levels = levels)
  }
  
  categorical_df$row_id <- NULL
  x_categorical <- model.matrix(~ . - 1, data = categorical_df)
  
  x <- cbind(x_numeric, x_categorical)
  storage.mode(x) <- "double"
  x
}

preprocessor <- fit_preprocessor(training)

x_train <- transform_data(training, preprocessor)
x_validation <- transform_data(validation, preprocessor)
x_test <- transform_data(testing, preprocessor)

y_train <- as.numeric(training[[label_column]])
y_validation <- as.numeric(validation[[label_column]])
y_test <- as.numeric(testing[[label_column]])

dim(x_train)
dim(x_validation)
dim(x_test)
```

## Class weights

The `PitNextLap = 1` class is less common, so the model can get high accuracy by predicting mostly `0`.

Class weights make mistakes on the smaller class count more during training.

```{r}
class_counts <- table(factor(y_train, levels = c(0, 1)))

if (any(class_counts == 0)) {
  class_weights <- NULL
  print("One class is missing from the training split, so class weights were not used.")
} else {
  class_weights <- as.list(sum(class_counts) / (2 * as.numeric(class_counts)))
  names(class_weights) <- c("0", "1")
}

class_counts
class_weights
```

## Build neural network models

This project tries three simple approaches:

1. A baseline neural network with one hidden layer.
2. A deeper neural network with two hidden layers.
3. A deeper neural network with dropout regularization.

Dropout is the regularization method. It randomly turns off part of the network during training so the model does not rely too heavily on one small pattern.

```{r}
build_model <- function(model_name, input_dim) {
  model <- keras_model_sequential()
  
  if (model_name == "baseline") {
    model %>%
      layer_dense(units = 64, activation = "relu", input_shape = c(input_dim)) %>%
      layer_dense(units = 1, activation = "sigmoid")
  }
  
  if (model_name == "deeper") {
    model %>%
      layer_dense(units = 128, activation = "relu", input_shape = c(input_dim)) %>%
      layer_dense(units = 64, activation = "relu") %>%
      layer_dense(units = 1, activation = "sigmoid")
  }
  
  if (model_name == "dropout_regularized") {
    model %>%
      layer_dense(units = 128, activation = "relu", input_shape = c(input_dim)) %>%
      layer_dropout(rate = 0.30) %>%
      layer_dense(units = 64, activation = "relu") %>%
      layer_dropout(rate = 0.30) %>%
      layer_dense(units = 1, activation = "sigmoid")
  }
  
  model %>% compile(
    loss = "binary_crossentropy",
    optimizer = "adam",
    metrics = "accuracy"
  )
  
  model
}
```

## Train and evaluate helper functions

Accuracy tells how often the model is right overall.

Precision answers: when the model predicts pit next lap, how often is it right?

Recall answers: out of the real pit-next-lap cases, how many did the model catch?

F1 balances precision and recall.

```{r}
calculate_metrics <- function(actual, predicted) {
  actual <- as.integer(actual)
  predicted <- as.integer(predicted)
  
  tp <- sum(actual == 1 & predicted == 1)
  tn <- sum(actual == 0 & predicted == 0)
  fp <- sum(actual == 0 & predicted == 1)
  fn <- sum(actual == 1 & predicted == 0)
  
  accuracy <- (tp + tn) / length(actual)
  precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
  recall <- ifelse(tp + fn == 0, 0, tp / (tp + fn))
  f1 <- ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall))
  
  list(
    accuracy = accuracy,
    precision = precision,
    recall = recall,
    f1 = f1,
    confusion_matrix = matrix(c(tn, fp, fn, tp), nrow = 2, byrow = TRUE)
  )
}

metric_value <- function(eval_result, metric_name, position) {
  if (!is.null(names(eval_result)) && metric_name %in% names(eval_result)) {
    return(as.numeric(eval_result[[metric_name]]))
  }
  as.numeric(eval_result[[position]])
}

evaluate_trained_model <- function(model_name, model) {
  train_eval <- evaluate(model, x_train, y_train, verbose = 0)
  validation_eval <- evaluate(model, x_validation, y_validation, verbose = 0)
  test_eval <- evaluate(model, x_test, y_test, verbose = 0)
  
  probabilities <- as.numeric(predict(model, x_test, verbose = 0))
  predictions <- ifelse(probabilities >= 0.5, 1, 0)
  metrics <- calculate_metrics(y_test, predictions)
  
  list(
    row = data.frame(
      model = model_name,
      train_loss = metric_value(train_eval, "loss", 1),
      validation_loss = metric_value(validation_eval, "loss", 1),
      test_loss = metric_value(test_eval, "loss", 1),
      train_accuracy = metric_value(train_eval, "accuracy", 2),
      validation_accuracy = metric_value(validation_eval, "accuracy", 2),
      test_accuracy = metric_value(test_eval, "accuracy", 2),
      precision = metrics$precision,
      recall = metrics$recall,
      f1 = metrics$f1
    ),
    probabilities = probabilities,
    predictions = predictions,
    confusion_matrix = metrics$confusion_matrix
  )
}
```

## Train the models

For a quick test, this uses fewer epochs. Increase `epochs_to_run` for the final experiment.

```{r}
epochs_to_run <- ifelse(quick_mode, 5, 20)
batch_size <- 256

model_names <- c("baseline", "deeper", "dropout_regularized")
trained_models <- list()
histories <- list()
all_results <- list()

for (model_name in model_names) {
  cat("\nTraining", model_name, "\n")
  
  model <- build_model(model_name, input_dim = ncol(x_train))
  
  summary(model)
  
  history <- fit(
    model,
    x = x_train,
    y = y_train,
    validation_data = list(x_validation, y_validation),
    epochs = epochs_to_run,
    batch_size = batch_size,
    class_weight = class_weights,
    verbose = 2
  )
  
  trained_models[[model_name]] <- model
  histories[[model_name]] <- history
  all_results[[model_name]] <- evaluate_trained_model(model_name, model)
}
```

## Compare model results

```{r}
metrics_summary <- do.call(rbind, lapply(all_results, function(result) result$row))
rownames(metrics_summary) <- NULL
metrics_summary
```

## Pick the best model

The best model is chosen by validation loss. Validation loss is used because the validation set helps compare models before checking the final test set.

```{r}
best_index <- which.min(metrics_summary$validation_loss)
best_model_name <- metrics_summary$model[best_index]
best_model_name

best_result <- all_results[[best_model_name]]
```

## Confusion matrix for the best model

Rows are actual values. Columns are predicted values.

```{r}
confusion <- best_result$confusion_matrix
rownames(confusion) <- c("actual_0_no_pit", "actual_1_pit")
colnames(confusion) <- c("predicted_0_no_pit", "predicted_1_pit")
confusion
```

## Plot training history

```{r}
plot_history <- function(history, model_name) {
  history_df <- as.data.frame(history$metrics)
  history_df$epoch <- seq_len(nrow(history_df))
  
  plot(
    history_df$epoch,
    history_df$loss,
    type = "l",
    xlab = "Epoch",
    ylab = "Loss",
    main = paste(model_name, "loss")
  )
  lines(history_df$epoch, history_df$val_loss, lty = 2)
  legend("topright", legend = c("training", "validation"), lty = c(1, 2))
  
  plot(
    history_df$epoch,
    history_df$accuracy,
    type = "l",
    xlab = "Epoch",
    ylab = "Accuracy",
    main = paste(model_name, "accuracy")
  )
  lines(history_df$epoch, history_df$val_accuracy, lty = 2)
  legend("bottomright", legend = c("training", "validation"), lty = c(1, 2))
}

plot_history(histories[[best_model_name]], best_model_name)
```

## View sample predictions

The model outputs a probability between `0` and `1`.

A probability greater than or equal to `0.5` is converted into prediction `1`.

```{r}
preview_count <- min(20, length(y_test))

prediction_preview <- data.frame(
  actual = y_test[1:preview_count],
  predicted_probability = round(best_result$probabilities[1:preview_count], 4),
  predicted_class = best_result$predictions[1:preview_count]
)

prediction_preview
```

## Save simple results

```{r}
dir.create("results", showWarnings = FALSE)

write.csv(metrics_summary, file.path("results", "simple_rmd_metrics_summary.csv"), row.names = FALSE)
saveRDS(preprocessor, file.path("results", "simple_rmd_preprocessor.rds"))

metrics_summary
```

## Final notes

This simplified version keeps the project focused on the required neural-network workflow:

- Load the dataset.
- Explain and choose features.
- Preprocess categorical and numeric data.
- Split into training, validation, and testing data.
- Train multiple neural networks.
- Include dropout regularization.
- Evaluate the models with accuracy, precision, recall, F1, and a confusion matrix.
