---
title: "F1 Pit Strategy"
author: "Blake Thompson"
date: "4/27/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## F1 example

This example uses a Formula 1 dataset to predict whether a driver will pit on the next lap.

```{r}
library(keras3)
library(tfdatasets)
library(tensorflow)
```

## dataframes

This chunk loads the dataset and shows the basic structure of the data.

```{r}
# Loading a csv file
f1df <- read.csv("data/f1_strategy_dataset_v4.csv")

# Using a smaller part of the data keeps the example easier to run while still training and testing the model.
f1df <- f1df[1:10000,]

# Various syntax for R dataframes
summary(f1df)

head(f1df)

# You can refer to columns by name or by number:
head(f1df[,1])

head(f1df$LapNumber)

head(f1df["LapNumber"])
```

```{r}
# You can refer to rows by numbers, note difference from columns:
f1df[5,]

# Ranges of columns:
head(f1df[,1:3])

# Ranges of rows:
f1df[1:3,]
```

## Data preprocessing

This chunk keeps simple numeric columns and keeps PitNextLap as the category the model is trying to predict.

```{r}
# Keeping only numeric columns for this simple version.
f1df <- f1df[,c(2,4,5,6,7,9,10,11,14,15,16,13)]

# plotting PitNextLap vs LapNumber
plot(f1df$LapNumber, f1df$PitNextLap, xlab = "Lap number", ylab = "Pit next lap")

head(f1df)
```

## Splitting data into training and testing

This chunk randomly separates the rows into training data and testing data.

```{r}
sample_size <- 8000
set.seed(1234)

picked <- sample(seq_len(nrow(f1df)),size = sample_size)
training <- f1df[picked,]
testing <- f1df[-picked,]

# Changing y into categorical data.
yTr <- to_categorical(training$PitNextLap)
yTest <- to_categorical(testing$PitNextLap)

yTest
```

## Neural network for the F1 example

This chunk creates one simple neural network that takes the numeric F1 columns as input and predicts the PitNextLap category.

```{r}
model = keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = c(11)) %>%
  layer_dense(units = ncol(yTr), activation = "softmax")

model %>% compile(
  loss = "categorical_crossentropy",
  optimizer = "adam", 
  metrics = "accuracy"
)

xTr <- as.matrix(training[,1:11])
xTest <- as.matrix(testing[,1:11])

model %>% 
  fit(
    x = xTr,
    y = yTr,
    epochs = 5
  )
```

# Evaluate the model

This chunk tests the model on data it did not train on and then shows the predicted probabilities.

```{r}
model %>% evaluate(xTest, yTest)

# Predicting likelihood of all categories:
result <- model %>% predict(xTest)

result

testing[,12]
```
